---
title: "example_qPCR_analysis"
author: "Libby Murphy"
date: "`Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Example qPCR analysis

This is an example qRT-PCR analysis using the delta-delta-Cq method. I normalized to a housekeeper gene and a calibrator. All genes and samples have been re-labeled to protect unpublished findings, so please do not consider this document to reflect any real-world biology.

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(purrr)
library(conflicted)
conflicts_prefer(dplyr::filter)
library(ggprism)

experiment <- "qPCR"
```

## QC comparisons 

```{r qc, echo=FALSE, message=FALSE, warning=FALSE}
# read in data file 
data <- read_csv(here::here(experiment, "data", "example_qPCR.csv")) %>% dplyr::select("Well", "Target", "Sample", "Cq")

Cq_QC <- ggplot(data = data, aes(y = Cq, x = Target, color = Sample)) +
  geom_jitter(width = .2, size = 2) + 
  theme_prism() + 
  ylim(0,40) +
  ggtitle("Cq") +
  theme(axis.text.x = element_text(angle = 60, hjust=0.5))

Cq_QC

# This looks good. I don't see any NTC wells crossing the threshold and replicates are grouping together.

# I have two housekeeper genes in this experiment: GENE6 and GENE7. I'll check the sd on both and normalize to whichever is tighter. The other will be discarded.
paste("sd GENE6 = ", (data[!is.na(data$Cq),] %>% dplyr::filter(Target == "GENE6", Sample != "CALIB_NA", Sample != "NTC_NA") %>% pull(Cq) %>% sd()), sep = "")
paste("sd GENE7 = ", (data[!is.na(data$Cq),] %>% dplyr::filter(Target == "GENE7", Sample != "CALIB_NA", Sample != "NTC_NA") %>% pull(Cq) %>% sd()), sep = "") # winner!

# remove H2ONTC and noRT samples
data_filt <- data %>% filter(Sample != "NTC_NA")

# remove all rows with a Cq of NaN
data_filt <- na.omit(data_filt)
```

QC plot looks nice. GENE7 was the superior HKG. Proceed with normalization

```{r normalization, echo=FALSE, message=FALSE, warning=FALSE}
# group data into individual conditions and calculate the condition mean. 
dataSum <- data_filt %>% 
  group_by(Target, Sample) %>%
  dplyr::summarise(Cq = mean(Cq)) %>%
  pivot_wider(names_from = Target, values_from = Cq)

# calculate deltaCq (dCq). Select other targets and subtract HKG from them to get dCq
data_dCq <- bind_cols(
  dataSum %>% dplyr::select(Sample),
  dataSum %>% purrr::keep(is.numeric) - dataSum$GENE7
  ) %>% 
  dplyr::select(-"GENE7")

tmp <- data_dCq %>% purrr::keep(is.numeric) %>% t() %>% as.data.frame()

colnames(tmp) <- data_dCq %>% dplyr::select(Sample) %>% as.matrix()

ddCq <- tmp %>%
  mutate(veh_compoundStim = veh_compoundStim - CALIB_NA) %>%
  mutate(veh_compoundVeh = veh_compoundVeh - CALIB_NA) %>%
  mutate(tx_compoundStim = tx_compoundStim - CALIB_NA) %>%
  mutate(tx_compoundVeh = tx_compoundVeh - CALIB_NA) %>%
  select(-CALIB_NA)

ddCq_lfc <- 2^-(ddCq)

ddCq_preplot <- data.frame(t(ddCq_lfc)) %>% rownames_to_column(var = "Sample") 

ddCq_plot <- reshape2::melt(ddCq_preplot, value.name = "ddCq", variable.name = "target") %>% separate(col = Sample, into = c("cocktail", "tx"), sep = "_")
```

```{r plotting, echo = FALSE, message=FALSE, warning=FALSE}
ddCq_plot$cocktail <- factor(ddCq_plot$cocktail, levels = c("veh", "tx"))
ddCq_plot$tx <- factor(ddCq_plot$tx, levels = c("compoundVeh", "compoundStim"))

pddCq <- ggplot(ddCq_plot[!c(grepl(pattern = "GENE6", ddCq_plot$target)),]) +
  geom_col(aes(x = cocktail, y = ddCq, fill = tx), position = position_dodge(), color = "black") +
  facet_grid(~ target) +
  theme_prism() + # I use ggprism for visual continuity with lab members who prefer Graphpad Prism.
  ylab("lfc") +
  scale_fill_manual(values = c("grey25", "grey75")) +
  theme(axis.text.x = element_text(angle = 60, hjust=0.5)) +
  ggtitle("example_qPCR")

pddCq
```

```{r save plots, echo = FALSE, message=FALSE, warning=FALSE}
ggsave(filename = "example_qPCR.png", plot = cowplot::plot_grid(Cq_QC, pddCq, nrow = 1), device = "png", height = 5, width = 15, path = here::here(experiment, "results"))

ggsave(filename = "example_qPCR_lfc.png", plot = pddCq, device = "png", height = 5, width = 10, path = here::here(experiment, "results"))

ggsave(filename = "example_qPCR_Cq.png", plot = Cq_QC, device = "png", height = 5, width = 8, path = here::here(experiment, "results"))

write_csv(ddCq_plot, file = here(experiment, "results", "example_qPCR_lfc.csv"))
```

The treatment induced an increase in all genes with compoundVeh. compoundStim strikingly ameliorated the increase in expression of all genes.